---
title: "Customization and advanced usage"
author: "Yiluan Song"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(batchplanet)

# Read the data file with coordinates
df_coordinates <- read_csv(system.file("extdata", "example_neon_coordinates.csv", package = "batchplanet"), show_col_types = FALSE)
head(df_coordinates)

# Set the parameters for the PlanetScope API
setting <- set_planetscope_parameters(
  api_key = set_api_key(),
  item_name = "PSScene",
  asset = "ortho_analytic_4b_sr",
  product_bundle = "analytic_sr_udm2",
  cloud_lim = 1,
  harmonized = T
)

# Set the data directory
# You should use `set_data_directory()` to set the data directory, but for demonstration purpose, we use the package's extdata folder.
dir_data <- system.file("extdata", package = "batchplanet")

# Set the bounding box for the site of interest
bbox <- set_bbox(df_coordinates, "SJER")
```

We understand that you may want to have more control over your workflow. Here are some examples of how you can use the functions in the package to customize your workflow.

## Direct search and order for a single site

If you would like to customize your batch downloading process, you may use the functions `search_planetscope_imagery()` and `order_planetscope_imagery()` to search and order images for a single site and time period. Be careful not to order too many images at once, as the Planet API has an internal limit. The `order_planetscope_imagery_batch()` function is designed to handle multiple sites and years in a more streamlined manner.
```{r, eval = F}
# Set the bounding box for the site of interest
bbox <- set_bbox(df_coordinates, "SJER")

images <- search_planetscope_imagery(
  api_key = setting$api_key,
  bbox = bbox,
  date_end = "2025-03-31",
  date_start = "2025-03-01",
  item_name = setting$item_name,
  asset = setting$asset,
  cloud_lim = setting$cloud_lim
)
order_id <- order_planetscope_imagery(
  api_key = setting$api_key,
  bbox = bbox,
  items = images,
  item_name = setting$item_name,
  product_bundle = setting$product_bundle,
  harmonized = setting$harmonized,
  order_name = "SJER_2025_60_90"
)
```

## Direct download from a single order

If you would like to customize your batch downloading process, you may use the function `download_planetscope_imagery()` to download images from a single order. You can find the order ID saved in the `raw/` folder, or you can find it in the past orders in your Planet account, if you did not use `order_planetscope_imagery_batch()` to order images.
```{r, eval = F}
download_planetscope_imagery(
  order_id = order_id,
  exportfolder = file.path(dir_data, "raw", "SJER", "SJER_2025_60_90"),
  api_key = setting$api_key,
  overwrite = F
)
```

## Direct retrieval of time series for a set of coordinates

If you would like to customize your time series retrieval process, you may use the function `retrieve_planetscope_time_series()` for a single folder `dir_site` and a set of coordinates you specify. 
```{r, eval = F}
df_coordinates_example <- df_coordinates %>% filter(site == "SJER", group == "Quercus")
df_ts_example <- retrieve_planetscope_time_series(
  dir_site = file.path(dir_data, "raw", "SJER"),
  sf_coordinates = df_coordinates_example %>% sf::st_as_sf(coords = c("lon", "lat"), crs = sf::st_crs(4326)),
  num_cores = 10
)
```

## Clean time series data in a single data frame

To process a single set of time series in a data frame, you can use the `clean_planetscope_time_series()` function.
```{r}
df_ts_example <- read_data_product(dir = dir_data, v_site = "SJER", v_group = "Quercus", product_type = "ts")
df_clean_example <- clean_planetscope_time_series(df_ts = df_ts_example, calculate_evi = T)
```

## Calculate phenological metrics from a single data frame

You can customize your own phenological metrics calculation, not necessarily PlanetScope data or even EVI data, using the `calculate_phenological_metrics()` function that takes a data frame of time series data and a data frame of thresholds and returns a data frame with the calculated phenological metrics.

```{r}
df_clean_example <- read_data_product(dir = dir_data, v_site = "SJER", v_group = "Quercus", product_type = "clean")
df_thres <- set_thresholds(thres_up = c(0.3, 0.4, 0.5), thres_down = NULL)
df_doy_example <- calculate_phenological_metrics(df_index = df_clean_example, df_thres = df_thres, var_index = "evi", min_days = 20, check_seasonality = F)
```