---
title: "Miscellaneous time series processing tools"
author: "Yiluan Song"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(batchplanet)
```

This package also provides tools for time series processing. These tools are not specific to PlanetScope data and can be used with any time series data, especially those with seasonality.

## Gap-filling and smoothing

We set up a simulated time series with a double logistic function, which resembles the greenness of deciduous trees. We add noise to the time series and introduce some missing data.
```{r}
t <- 1:365

# Double logistic function: leaf-on and leaf-off phases
double_logistic <- function(t, L = 0.1, U = 0.6, k1 = 0.1, k2 = 0.1, t1 = 120, t2 = 280) {
  L + (U - L) * (1 / (1 + exp(-k1 * (t - t1)))) * (1 / (1 + exp(k2 * (t - t2))))
}

# Generate simulate_ts values using double logistic
simulate_ts <- double_logistic(t)

# Add Gaussian noise
set.seed(42)
simulate_ts <- simulate_ts + rnorm(length(t), mean = 0, sd = 0.1)

# Introduce missing data (e.g., cloud cover)
missing_idx <- sample(1:length(t), size = round(0.2 * length(t))) # 10% missing
simulate_ts[missing_idx] <- NA
```

The `whittaker_smoothing_filling()` function is used to smooth the time series and fill in missing values within one step. Under the hood, it uses a weighted Whittaker smoothing (the `whit1` funtion from the `ptw` package), but we assign zero weight to any missing value and equal weight to all other values, such that the missing values are filled with the smoothed values.

The `lambda` parameter controls the smoothness of the output. A larger value results in a smoother output, while a smaller value retains more of the original signal. It is recommended to try different values and visualize the time series before and after smoothing. The `maxgap` parameter specifies the maximum gap size for filling missing values. Any consecutive missing values larger than `maxgap` will not be filled. The `minseg` parameter specifies the minimum segment length for smoothing. Segments shorter than this length will be treated as missing and then either left missing or filled with smoothed values, depending on `maxgap`. This option allows us to avoid smoothing over very short segments of data that may not be representative of the overall trend.

```{r}
smoothed_ts <- whittaker_smoothing_filling(
  x = simulate_ts,
  lambda = 50,
  maxgap = 30,
  minseg = 2
)

# Compare original and smoothed time series
df_compare <- data.frame(
  original = simulate_ts,
  smoothed = smoothed_ts
) %>%
  mutate(timestep = row_number())

ggplot(df_compare) +
  geom_point(aes(x = timestep, y = original, color = "original")) +
  geom_line(aes(x = timestep, y = smoothed, color = "smoothed"), linewidth = 2, alpha = 0.75) +
  scale_color_manual(values = c("original" = "black", "smoothed" = "dark green")) +
  theme_minimal() +
  labs(
    x = "Timestep",
    y = "Value",
    color = ""
  ) +
  theme(legend.position = "bottom")
```

## Determine seasonality

The `determine_seasonality()` function is used to check if a time series from one year (or one life cycle) has seasonal variations or has no significant variations. In practice, we can use this function to distinguish if the greenness of a tree within a year has seasonal variations (e.g., deciduous trees) or remains relatively constant (e.g., evergreen trees). Under the hood, the function fits a simple linear regression model and segmented regression models with 1 to 3 breakpoints to a given time series. It then compares their Akaike Information Criterion (AIC) values (with a penalty parameter k) to assess whether a simple linear regression (i.e., non-segmented) model is preferred. The `k` parameter controls the penalty for the number of breakpoints in the calculation of AIC. A larger `k` means that the favored model is more likely to be a simple linear regression model, such that we are more conservative in concluding that the time series has seasonal variations.

```{r}
# Check if a time series is flat
example_seasonality <- determine_seasonality(
  ts = simulate_ts,
  k = 50
)

print(str_c("Time series has significant seasonal variation: ", example_seasonality))
```
