---
title: "Vignette"
author: "Yiluan Song, Ken Reid"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This workflow demonstrates the features of the _batchplanet_ package. It provides a step-by-step guide to downloading PlanetScope imagery, processing time series data, and calculating phenological metrics. The package is designed to work with the PlanetScope API and is particularly useful for researchers and practitioners in environmental studies and remote sensing.

Who should use this package?
- If you prioritize reproducibility and control over data.
- If you prefer using R to python.
- If you want to download PlanetScope imagery over multiple sites and over a long time (optional but recommended).
- If you use high-performance computing (optional).
- If you want to retrieve and process time series from PlanetScope data (optional).

```{r}
library(batchplanet)
```

## 1 Read coordinate data
In order to download PlanetScope imagery, you need to provide coordinates of points of interest. For this example, we will use a sample data file retrieved from the [National Ecological Observatory Network (NEON)](https://www.neonscience.org/) [plant phenology observations data product](https://data.neonscience.org/data-products/DP1.10055.001).
```{r}
# Read the data file with coordinates
df_coordinates <- read_csv(system.file("extdata", "example_neon_coordinates.csv", package = "batchplanet"), show_col_types = FALSE)
head(df_coordinates)
```
You should create your own data frame with coordinates. Make sure your data frame for coordinates have the columns: `id` (unique id for each point of interest), `lon` for longitude, and `lat` for latitude. You may have a character column `site` if you have clustered coordinates at dispersed sites. You may also a character column `group` you can use to group coordinates later on.

What do we mean by "clustered coordinates at dispersed sites?" Use the function `visualize_coordinates()` to visualize the coordinates. Our sample coordinates are dispersed across 47 NEON sites. Within each NEON site, there are above 200 coordinates in a square, representing individual trees tagged for phenology observations.
```{r}
visualize_coordinates(df_coordinates) # Multiple sites across continental US
visualize_coordinates(df_coordinates %>% filter(site == "SJER")) # Zoom in to one site SJER
```
Why is it important to label the coordinates with sites? If we try to download PlanetScope imagery that cover all coordinates, we will end up with a tremendous amount of imagery that cover most of continental US. Instead, we will use the site labels to create bounding boxes that cover each site. This way, we can download imagery that only cover the sites of interest. If your coordinates are already clustered in a small area, you don't have to supply the site labels. The package will process all coordinates as a single site.

## 2 Download PS data

You will need an active Planet account and an API key to access the PlanetScope API. You can sign up for an account on the [Planet website](https://www.planet.com/get-started/). Once you have an account, you can copy your API key from your account settings.

### 2.1 Set downloading parameters

You will specify several key parameters for downloading PlanetScope imagery using the `set_planetscope_parameters()` function. These parameters include the API key, item name, asset type, product bundle, cloud cover limit, and whether to use harmonized data. 

Within this step, you set your API key using the `set_api_key()` function. This function will save your API key in a hidden environment file in your working directory, so you don't have to enter it every time you run the code. If you would like to change your API key, you can use the same function with the argument `change_key = T`.

>In the Data API, an __item__ is an entry in our catalog, and generally represents a single logical observation (or scene) captured by a satellite. Items have __assets__, which are the downloadable products derived from the item's source data. [PlanetScope documentation](https://docs.planet.com/develop/apis/data/items/#assets)

> __Product bundles__ comprise of a group of assets for an item. [PlanetScope documentation](https://docs.planet.com/develop/apis/orders/product_bundles/)

Our default item is `PSScene`, 8-band PlanetScope imagery. The default asset is `ortho_analytic_4b_sr`, which is a PlanetScope atmospherically corrected surface reflectance product with four bands (blue, green, red, near-infrared). The default product bundle is `analytic_sr_udm2`, which includes the surface reflectance data and the [Usable data mask 2 (UDM2)](https://docs.planet.com/data/imagery/udm/). You can change these parameters to suit your needs. See a full list of PlanetScope items and assets in the [PlanetScope data catalog](https://docs.planet.com/data/) and see information and product bundles [here](https://docs.planet.com/develop/apis/orders/product_bundles/). 

PlanetScope API allows filtering imagery by several criteria. Here, we allow users to filter imagery below a certain cloud coverage (the ratio of the area covered by clouds to total area). The `cloud_lim` parameter is set to 1 by default, which means we will download all images regardless of cloud coverage. You can set this parameter to a lower value (e.g., 0.5) to filter out images with more than 50% cloud coverage.

PlanetScope API allow users to apply a tool named "harmonize" that applies scene-level normalization and harmonization, such that all PlanetScape data were consistent and approximately comparable to data from Sentinel 2. This tool is useful when integrating or comparing images from different times and locations. Refer to [PlanetScope technical documentation](https://assets.planet.com/docs/scene_level_normalization_of_planet_dove_imagery.pdf) for more details. The `harmonized` parameter is set to `TRUE` by default, which means we will download harmonized data. You can set this parameter to `FALSE` to download non-harmonized data.

```{r}
setting <- set_planetscope_parameters(
  api_key = set_api_key(),
  item_name = "PSScene",
  asset = "ortho_analytic_4b_sr",
  product_bundle = "analytic_sr_udm2",
  cloud_lim = 1,
  harmonized = T
)
```

Use the `set_data_directory()` function to specify a directory to store the downloaded data and later store processed data. It is recommended to use symbolic link to a directory on your HPC system. This way, you can easily access the data from your local machine without being limited by the storage space on your local machine. Please still closely monitor the storage space on your local machine or HPC system as the data can be large. For example, all images for New York City since 2017 can be over 1TB.
```{r}
dir_data <- set_data_directory()
```

### 2.2 Order

Use the `order_planetscope_imagery_batch()` function to order PlanetScope imagery over multiple sites and years. This function first searches for all the images that meets the criteria you specified, and then orders the images.

You can specify the sites (must exist in your `site` column of the coordinates data frame) and years (after 2014) you want to order images for. Here, we order images from two sites in one year as an example. If you do not specify `v_site` and `v_year`, the function will order images for all sites and years in the coordinates data frame.

The function will create a folder `raw/` in the data directory you specified, and multiple subfolders for different sites you specified. In each subfolder, the function will save the order IDs in rds files for later use.

```{r, eval = F}
order_planetscope_imagery_batch(dir = dir_data, df_coordinates = df_coordinates, v_site = c("HARV", "SJER"), v_year = 2025, setting = setting)
```
If you would like to customize your batch downloading process, you may use the functions `search_planetscope_imagery()` and `order_planetscope_imagery()` to search and order images for a single site and time period. Be careful not to order too many images at once, as the Planet API has an internal limit. The `order_planetscope_imagery_batch()` function is designed to handle multiple sites and years in a more streamlined manner.

After successful ordering, check your [Planet account](https://www.planet.com/account/) to make sure orders are completed. Make sure all orders are "success" without any order "failed." Once all orders are completed, you might proceed to the next step of downloading. Please do not order the same images repeatedly.

> Why would you see failed orders?
> - You have exceeded your Planet account limit. If you have exceeded your limit, you can either wait for the limit to reset or contact Planet support to increase your limit.
> - The order is too large. You can try to order images for a smaller area or a shorter time period.
> - There were errors in the query, such as invalid coordinates. You can preview the images in your Planet Account.
> - Sometimes an order may fail due to temporary issues with the Planet API. In this case, you can try reordering the images after a few hours or days. Instead of reordering the entire batch, you can reorder only the failed sites and years.

### 2.3 Download

Use the `download_planetscope_imagery_batch()` function to download the ordered images. This function will save downloaded images to the folder `raw/` in the data directory. If you do not specify `v_site` and `v_year`, the function will download all ordered images in the `raw/` folder. This step uses multiple cores to download. Use `num_cores = 12` to download data from 12 months in parallel. `overwrite = F` prevents overwriting existing images.

```{r, eval = F}
download_planetscope_imagery_batch(dir = dir_data, v_site = c("HARV", "SJER"), v_year = 2025, setting = setting, num_cores = 3, overwrite = F)
```

If you would like to customize your batch downloading process, you may use the function `download_planetscope_imagery()` to download images from a single order. You can find the order ID saved in the `raw/` folder, or you can find it in the past orders in your Planet account, if you did not use `order_planetscope_imagery_batch()` to order images.

At this point, you might have downloaded a large amount of images. Consider archiving these raw images and removing the local copy after you have completed your downstream analyses.

### 2.4 Visualize true color imagery

The `visualize_true_color_imagery_batch()` function starts a shiny app to visualize multiple images in the data directory. In the app, you can then choose the site and time you want to visualize and toggle brightness. Again, you have the option to overlay the coordinates on the images.
```{r, eval = F}
visualize_true_color_imagery_batch(dir = dir_data, df_coordinates = df_coordinates)
```

You can visualize the true color imagery of single images using the `visualize_true_color_imagery()` function. You can overlay the coordinates from the relevant site on the imagery (optional). You can specify the brightness of the image using the `brightness` parameter.
```{r}
df_coordinates_SJER <- df_coordinates %>% filter(site == "SJER")

visualize_true_color_imagery(
  file = system.file("extdata", "raw/SJER/SJER_2025_60_90/20250318_190514_92_24d3_3B_AnalyticMS_SR_harmonized_clip.tif", package = "batchplanet"),
  df_coordinates = df_coordinates_SJER,
  brightness = 5
)
```

## 3 Retrieve and process time series

### 3.1 Retrieve time series

The `retrieve_planetscope_time_series_batch()` function retrieves time series data from the downloaded PlanetScope imagery. This function uses coordinates from the `df_coordinates` data frame to extract reflectance values across all images at the relevant site.

The `v_site` and `v_group` parameters allow you to specify the sites and groups of coordinates you want to extract time series data for. Here, we extract time series for two sites (HARV and SJER) and trees from two genera (_Acer_ spp. and _Quercus_ spp.). The `max_sample` parameter limits the number of samples to be extracted from each site and group. When the number of coordinates is larger than `max_sample`, the function will randomly sample coordinates to extract time series data. The grouping option and `max_sample` are useful when you have a large number of coordinates and want to limit the computing time. We recommend limiting the number of coordinates per group per site to around 2000. If you do not specify `v_site` and `v_group`, the function will extract time series data for all sites available and treat all coordinates as one group.

The `num_cores` parameter allows you to use multiple cores for this step, which can significantly speed up the process. Use as many as you can afford, as each core reads a map layer in parallel.

This function will create a folder `ts/` in the data directory you specified and save time series data with files names in the format of `ts_<site>_<group>.rds`.

```{r, eval = F}
retrieve_planetscope_time_series_batch(dir = dir_data, df_coordinates = df_coordinates, v_site = c("HARV", "SJER"), v_group = c("Acer", "Quercus"), max_sample = 2000, num_cores = 10)
```

You can read in processed data using the `read_data_product()` function. To read in time series retrieved from raw imagery, set `product_type = "ts"`. The function will read in all time series data from the `ts/` folder and combine them into one data frame. You can specify the sites and groups you want to read in using the `v_site` and `v_group` parameters. If you do not specify these parameters, the function will read in all time series data.

You can visualize time series data using the `visualize_time_series()` function. You can specify the variable you want to plot, which needs to be a column in the supplies data frame, and the corresponding y-axis label. The optional `facet_var` parameter allows you to specify the variable you want to use for faceting the plot (e.g., site, group, or id). The `smooth` parameter allows you to specify whether you want to smooth the time series data or not. If `smooth = TRUE`, the function will use a weighted Whittaker smoothing method to smooth and fill the time series data (see Section 4.1). This function is not specific to PlanetScope data and can be used with any time series data, as long as the data frame has `date` or `time` at regular intervals, as well as the variable you want to plot.
```{r}
df_ts <- read_data_product(dir = dir_data, v_site = c("HARV", "SJER"), v_group = c("Acer", "Quercus"), product_type = "ts")
visualize_time_series(df_ts, var = "green", ylab = "Green reflectance", facet_var = "site", smooth = F)
```

### 3.2 Clean time series

The `clean_planetscope_time_series_batch()` function removes low quality data in the retrieved time series. We keep data points that meet all of the following criteria:
- The sun elevation angle is greater than 0 degrees (i.e., daytime images).
- The reflectance values for all bands (blue, green, red, and near-infrared) are greater than 0.
- The pixel was clear, had no snow, ice, shadow, haze, or cloud.
- The usable data mask had algorithmic confidence in classification ≥ 80% for the pixel.

The `calculate_evi` option allows you to calculate the Enhanced Vegetation Index (EVI) for the time series data and filter for EVI values between 0 and 1. The equation for EVI is as follows:
$$
EVI = 2.5 \times \frac{NIR - Red}{NIR + 6 \times Red - 7.5 \times Blue + 1}
$$
The function will create a folder `clean/` in the data directory you specified and save cleaned time series data with files names in the format of `clean_<site>_<group>.rds`.

```{r, eval = F}
clean_planetscope_time_series_batch(dir = dir_data, v_site = c("HARV", "SJER"), v_group = c("Acer", "Quercus"), num_cores = 3, calculate_evi = T)
```

To process a single set of time series in a data frame, you can use the `clean_planetscope_time_series()` function.

Similar to Section 3.2, you can read in cleaned time series data using the `read_data_product()` function, with `product_type = "clean"`. You can then visualize the cleaned time series data using the `visualize_time_series()` function. The cleaned time series data will have a new column `evi` that we can visualize, if you set `calculate_evi = T` in the previous step.

```{r}
df_clean <- read_data_product(dir = dir_data, v_site = c("HARV", "SJER"), v_group = c("Quercus"), product_type = "clean")
visualize_time_series(df_clean, var = "evi", ylab = "EVI", facet_var = "site", smooth = T)
```

### 3.3 Calculate and phenological metrics

When analyzing time series with seasonality, we often need to calculate the day of year when critical transitions occur. In the field of phenology, we are often interested in the time of 50% green-up and 50% green-down, which reflect the start and end of the growing season. We use the `calculate_phenological_metrics_batch()` function to calculate the day of year (DOY) for the increase and decrease of the specified index at specified thresholds.

> For individual trees at NEON sites monitored for phenology, we used the EVI time series to identify the green-up phases empirically. The end of a green-up phase (usually in the summer) was determined as the day of year when EVI reaches the maximum in the growing season. The start of a green-up phase (usually in the winter) was then determined as the day of year when EVI is at the minimum, prior to the end of the green-up phase. We then determined the timing of green-up at the 50% threshold (usually in the spring). This empirical method of defining green-up/down time has been widely applied to remote-sensing data in order to be compatible with different plant functional types with various seasonality that exhibit intra-annual changes in greenness. [Song et al., (2025)](https://doi.org/10.1016/j.srs.2025.100205)

`df_thres` is a data frame that specifies the thresholds for calculating phenological metrics. You can generate this data frame using the `set_thresholds()` function. The `thres_up` and `thres_down` columns can be set to `NULL` if you do not want to calculate phenological metrics for the increase or decrease of the specified index.

The `var_index` parameter specifies the variable you want to use for calculating phenological metrics (e.g., EVI).

The `min_days` parameter specifies the minimum number of days with valid data at a pixel in a life cycle, for phenological metrics to be estimated for the pixel in the life cycle. This is useful when you want to reduce the impacts of missing data points on the estimation of phenological metrics. Here, we set it to 20 days for demonstration as we demonstrate with data in the first three months of 2025, but when you have more than one year of data, you can set it to a larger number (e.g., 300 days) to ensure that the phenological metrics are estimated for pixels with sufficient data.

The `check_seasonality` parameter allows you to choose if you only extract phenological metrics for pixel and life cycles with significant seasonal variations (see Section 4.2). Here, we set it to `F` again as we do not demonstrate with a full year of data, but we recommend setting it to `T` when you would like to exclude pixels with no significant seasonal variations, such as some evergreen trees that show little seasonal variations in greenness, and get more reliable phenological metrics.

The `extend_to_previous_year` and `extend_to_next_year` parameters specify the number of days to extend the time series to include data from previous and next years, respectively.
> We extended the time series in each year from day 275 (Oct 2) in the previous calendar year to day 90 (Mar 30) in the following year (spanning 546 days) in order to include at least one full growing season with green-up and green-down. This step was necessary for the detection of green-up day when EVI increases from the minimum before the New Year, and the detection of green-down day when EVI decreases to the minimum after the New Year. [Song et al., (2025)](https://doi.org/10.1016/j.srs.2025.100205)

The function will create a folder `doy/` in the data directory you specified and save DOY data with files names in the format of `doy_<site>_<group>.rds`.

```{r}
df_thres <- set_thresholds(thres_up = c(0.3, 0.4, 0.5), thres_down = NULL)

calculate_phenological_metrics_batch(dir = dir_data, v_site = "SJER", v_group = "Quercus", df_thres = df_thres, var_index = "evi", min_days = 20, check_seasonality = F, extend_to_previous_year = 275, extend_to_next_year = 90, num_cores = 1)
```
This step is not specific to PlanetScope data. Neither it's specific to the EVI index. You can use this function to calculate phenological metrics for any time series data with seasonality. Make sure you format your data frame similar to the ones we use in the demonstration. You can find examples in the `inst/extdata/clean/` folder of the package. You can also customize your own workflow using the `calculate_phenological_metrics()` function that takes a data frame of time series data and a data frame of thresholds and returns a data frame with the calculated phenological metrics.

Similarly, we can read in phenological metrics data using the `read_data_product()` function, with `product_type = "doy"`. We can visualize the phenological metrics overlaid on the EVI time series data using the `visualize_time_series()` function.
```{r}
v_id <- c("NEON.PLA.D17.SJER.06001", "NEON.PLA.D17.SJER.06337", "NEON.PLA.D17.SJER.06310")
# Read and visualize the DOY results
df_doy_sample <- read_data_product(dir = dir_data, v_site = "SJER", v_group = "Quercus", product_type = "doy") %>%
  filter(id %in% v_id)

df_evi_sample <- read_data_product(dir = dir_data, v_site = "SJER", v_group = "Quercus", product_type = "clean") %>%
  filter(id %in% v_id)

visualize_time_series(df_ts = df_evi_sample, df_doy = df_doy_sample, var = "evi", ylab = "EVI", facet_var = "id", smooth = T)
```

## 4 Miscellaneous time series processing tools

This package also provides tools for time series processing. These tools are not specific to PlanetScope data and can be used with any time series data, especially those with seasonality.

We set up a simulated time series with a double logistic function, which resembles the greenness of deciduous trees. We add noise to the time series and introduce some missing data.
```{r}
t <- 1:365

# Double logistic function: leaf-on and leaf-off phases
double_logistic <- function(t, L = 0.1, U = 0.6, k1 = 0.1, k2 = 0.1, t1 = 120, t2 = 280) {
  L + (U - L) * (1 / (1 + exp(-k1 * (t - t1)))) * (1 / (1 + exp(k2 * (t - t2))))
}

# Generate simulate_ts values using double logistic
simulate_ts <- double_logistic(t)

# Add Gaussian noise
set.seed(42)
simulate_ts <- simulate_ts + rnorm(length(t), mean = 0, sd = 0.1)

# Introduce missing data (e.g., cloud cover)
missing_idx <- sample(1:length(t), size = round(0.2 * length(t))) # 10% missing
simulate_ts[missing_idx] <- NA
```

The `whittaker_smoothing_filling()` function is used to smooth the time series and fill in missing values within one step. Under the hood, it uses a weighted Whittaker smoothing (the `whit1` funtion from the `ptw` package), but we assign zero weight to any missing value and equal weight to all other values, such that the missing values are filled with the smoothed values.

The `lambda` parameter controls the smoothness of the output. A larger value results in a smoother output, while a smaller value retains more of the original signal. It is recommended to try different values and visualize the time series before and after smoothing. The `maxgap` parameter specifies the maximum gap size for filling missing values. Any consecutive missing values larger than `maxgap` will not be filled. The `minseg` parameter specifies the minimum segment length for smoothing. Segments shorter than this length will be treated as missing and then either left missing or filled with smoothed values, depending on `maxgap`. This option allows us to avoid smoothing over very short segments of data that may not be representative of the overall trend.

```{r}
smoothed_ts <- whittaker_smoothing_filling(
  x = simulate_ts,
  lambda = 50,
  maxgap = 30,
  minseg = 2
)

# Compare original and smoothed time series
df_compare <- data.frame(
  original = simulate_ts,
  smoothed = smoothed_ts
) %>%
  mutate(timestep = row_number())

ggplot(df_compare) +
  geom_point(aes(x = timestep, y = original, color = "original")) +
  geom_line(aes(x = timestep, y = smoothed, color = "smoothed"), linewidth = 2, alpha = 0.75) +
  scale_color_manual(values = c("original" = "black", "smoothed" = "dark green")) +
  theme_minimal() +
  labs(
    x = "Timestep",
    y = "Value",
    color = ""
  ) +
  theme(legend.position = "bottom")
```

The `determine_seasonality()` function is used to check if a time series from one year (or one life cycle) has seasonal variations or has no significant variations. In practice, we can use this function to distinguish if the greenness of a tree within a year has seasonal variations (e.g., deciduous trees) or remains relatively constant (e.g., evergreen trees). Under the hood, the function fits a simple linear regression model and segmented regression models with 1 to 3 breakpoints to a given time series. It then compares their Akaike Information Criterion (AIC) values (with a penalty parameter k) to assess whether a simple linear regression (i.e., non-segmented) model is preferred. The `k` parameter controls the penalty for the number of breakpoints in the calculation of AIC. A larger `k` means that the favored model is more likely to be a simple linear regression model, such that we are more conservative in concluding that the time series has seasonal variations.

```{r}
# Check if a time series is flat
example_seasonality <- determine_seasonality(
  ts = simulate_ts,
  k = 50
)

print(str_c("Time series has significant seasonal variation: ", example_flat_check))
```
