---
title: "workflow for juniper"
author: "Yiluan Song"
date: "2023-10-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(batchplanet)
```

## Read Juniper tree inventory

```{r}
df_tree <- misc_read_juniper_coord(version = 3)
vis_coord(df_tree)
```

Area too big, so divide them into multiple sites (counties).
```{r}
sp_county <- read_county(country = "US")
# sp_county_sub <- read_county_sub(country = "US", state = "texas") # county subdivision not used
df_tree_county <- util_coord_in_area(df_tree, sp_county, field = "NAME")
head(df_tree_county)
```

## Download PS data
Set some parameters. Replace API key with your own here.
```{r}
setting <- set_ps(api_key = "REMOVED", # songyl
                  harmonized = T)
setting
```

Order. Order IDs saved as rds.
```{r, eval = F}
order_ps_batch(dir = "alldata/juniper_ashei/PSdata/", df_plant = df_tree_county, v_site = NULL, setting = setting)
```

Here, check your Planet account to make sure orders are completed. Make sure all orders are "success" without any order "failed." Once all orders are completed, please do not order the same images repeatedly.

Download. Use Great Lakes R session with 12 cores for this step.
```{r, eval = F}
down_ps_batch(dir = "alldata/juniper_ashei/PSdata/", v_site = NULL, setting = setting)
```
Here, tar and compress folders for cities downloaded by submitting Slurm jobs and then move tarred compressed files to Data Den.

## Retrieve time series

Extract time series at tree locations. Use Great Lakes R session with 36 cores for this step.
```{r, eval = F}
proc_ps_ts(dir = "alldata/juniper_ashei/PSdata/", df_plant = df_tree_county, v_site = NULL, v_taxa = "Juniper ashei")
```

Time series preprocessing, including removing low quality data and calculating EVI.
```{r, eval = F}
proc_evi_ts(dir = "alldata/juniper_ashei/PSdata/", v_site = NULL, v_taxa = "Juniper ashei")
```

Read in data and save.
```{r}
df_index <- read_evi(dir = "alldata/juniper_ashei/PSdata/evi/", v_site = NULL, v_taxa = "Juniper ashei") %>%
  # select(-site) %>%
  mutate(r2g = red / green) %>%
  mutate(yellow = (red + green) / blue) %>%
  filter(yellow <= 10) # /(green/blue)*(red-blue+1))
df_index
df_index_version <- df_index %>% left_join(df_tree %>% distinct(id, version, group))
write_csv(df_index_version, "alldata/juniper_ashei/Jan24/indices.csv")
```

Calculate and visualize red to green ratio.
```{r}
vis_ts(df_index, var = "r2g", ylab = "R/G", smooth = T)
```

Calculate and visualize yellow.
```{r}
vis_ts(df_index, var = "yellow", ylab = "yellow", smooth = T)
```

The two chunks below applies to version 1 and version 2.
Any difference in the two versions? First version male trees, second version female trees.
```{r, eval = F}
vis_ts(df_index_version, n_id = 3, var = "r2g", ylab = "R/G", smooth = T, facet = "version")
vis_ts(df_index_version, n_id = 3, var = "yellow", ylab = "yellow", smooth = T, seed = NULL, facet = "version")
```

The chunk below applies to version 3.
Any difference in the two groups? First group green trees, second group orange trees.
```{r, eval = F}
vis_ts(df_index_version , v_site = "Burnet", v_year = 2021: 2024, n_id = 50, var = "r2g", ylab = "R/G", smooth = T, facet = "group")
vis_ts(df_index_version,  v_site = "Burnet",v_year = 2021: 2024, n_id = 50, var = "yellow", ylab = "yellow", smooth = T,facet = "group")
```